---
title: "Reproducible Research: Peer Assessment 1"
author: "Panos Rontogiannis"
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: true
---

Firstly let's set a few global *knitr* options and load the *lattice* package which will be used for plotting.


```{r global_options}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='figure/', warning = FALSE)
library(lattice) 
```

## Loading and preprocessing the data

Before loading the data we need to unzip it first. No preprocessing at this point. Later on we will add a couple collumns.

```{r loading_data}
unzip('activity.zip')
act <- read.csv('activity.csv')
```

## What is mean total number of steps taken per day?
```{r steps_per_day}
day_sum <- aggregate(steps ~ date, data = act, FUN = sum)

histogram(day_sum$steps, xlab = 'steps/day', main = 'Total number of steps taken each day', breaks = 15)

mean_steps_per_day <- mean(day_sum$steps, na.rm = TRUE)
median_steps_per_day <- median(day_sum$steps, na.rm = TRUE)

# Round values using sprintf. Somehow the round function didn't seem to work as i wanted it.
mean_steps_per_day <- sprintf("%.2f", mean_steps_per_day)
```

The **mean** of the total number of steps taken per day is `r mean_steps_per_day`.  
The **median** of the total number of steps taken per day is `r median_steps_per_day`.

## What is the average daily activity pattern?
```{r daily_pattern}
interval_mean <- aggregate(steps ~ interval, data = act, FUN = mean)

xyplot(steps ~ interval, data = interval_mean, type = 'l', main = 'Average number of steps taken per 5-minute interval', xlab = '5-minute interval')

max_interval <- which.max(interval_mean$steps)
```

The 5-minute interval, on average across all the days in the dataset, containing the maximum number of steps is interval **`r max_interval`**.

## Imputing missing values

```{r count_missing}
total_incomplete_rows <- length(which(!complete.cases(act)))
```

The total number of missing values in the dataset is **`r total_incomplete_rows`**.

To fill in the missing steps we use the mean for that 5-minute interval (one of the suggested strategies mentioned in the assignment description).

```{r fill_missing}
act_filled <- act

# I'm pretty sure this can be done in a functional style but haven't figured it out yet. Hence the for loop.
for (i in 1:nrow(act_filled)) {
    if (is.na(act_filled$steps[i])) {
        act_filled$steps[i] <- interval_mean$steps[interval_mean$interval == act_filled$interval[i]]
    }
}

day_sum_filled <- aggregate(steps ~ date, data = act_filled, FUN = sum)

histogram(day_sum_filled$steps, xlab = 'steps/day', main = 'Total number of steps taken each day', breaks = 15)

mean_steps_per_day_filled <- mean(day_sum_filled$steps)
median_steps_per_day_filled <- median(day_sum_filled$steps)

# Rounding.
mean_steps_per_day_filled <- sprintf("%.2f", mean_steps_per_day_filled)
median_steps_per_day_filled <- sprintf("%.2f", median_steps_per_day_filled)
```

After filling in the missing values, the **mean** of the total number of steps taken per day becomes `r mean_steps_per_day_filled` (from `r mean_steps_per_day`).  
The **median** of the total number of steps taken per day becomes `r median_steps_per_day_filled` (from `r median_steps_per_day`).

We can see that the mean has not changed but that the median has converged with the mean. They are now the same.

## Are there differences in activity patterns between weekdays and weekends?

First we add a column with the date in POSIXlt format. This is needed to identify the type of day using the *strftime* function. Then we add another column representing the type of day which can either be "weekdays" or "weekends".


```{r day_pattern_preparation}
act_filled$datePOSIX <- as.POSIXlt(act_filled$date, format = '%Y-%m-%d')
# Add a column with the type of day
act_filled$dayType <- ifelse(strftime(act_filled$datePOSIX, format = '%u') > 5, 1, 0)
# and convert it to a factor.
act_filled$dayType <- factor(act_filled$dayType, levels = c(0, 1), labels = c('weekday', 'weekend'))
```

To compute the mean intervals per type of day, we first subset the filled data into two dataframes. One per type of day. Then compute the mean per interval for both types of day. Then we combine these two dataframes into one. So in the combined dataframe there will be two rows per interval. One containing the mean steps for weekends and the other for weekdays. Finally we created the panel plot.

```{r day_pattern_compute}
# Subset according to type of day.
wdays <- subset(act_filled, dayType == "weekday")
wends <- subset(act_filled, dayType == "weekend")

# Calculate the mean steps per interval for each type of day.
wdays_mean <- aggregate(steps ~ interval, data = wdays, FUN = mean)
wends_mean <- aggregate(steps ~ interval, data = wends, FUN = mean)
# Then add the dayType. This is going to be used by xyplot to create the panel plot.
wdays_mean$dayType <- "weekday"
wends_mean$dayType <- "weekend"
# Finally  combine the two data frames into one that contain two means for each interval. One for weekdays and the other for weekends. I'm certain there is a more ellegant way but it works.
combined_mean <- rbind(wends_mean, wdays_mean)

xyplot(steps ~ interval | dayType, data = combined_mean, type='l', layout = c(1, 2), xlab = '5-minute interval', ylab = 'Number of steps')
```

From the panel plot we can see that on weekdays the mean steps seem higher betweem intervals 500 and 900 but from then on seem to be lower than on weekends. This can aslo be seen from summary statistics on the weekend and weekdays dataframes. weekdays have a higher range but lower median and mean.

```{r day_pattern_summary_stats}
summary(wends_mean)
summary(wdays_mean)
```

I don't know why though :)