---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
```{r}
unzip('activity.zip')
activity <- read.csv('activity.csv')
```

## What is mean total number of steps taken per day?
```{r}
day_sum <- aggregate(steps ~ date, data = activity, FUN = sum)

hist(day_sum$steps, xlab = 'steps/day', main = 'Total number of steps taken each day', breaks = 15)

mean_steps_per_day <- mean(day_sum$steps, na.rm = TRUE)
median_steps_per_day <- median(day_sum$steps, na.rm = TRUE)
```

The mean of the total number of steps taken per day is **`r mean_steps_per_day`**.  
The median of the total number of steps taken per day is **`r median_steps_per_day`**.

## What is the average daily activity pattern?
```{r}
interval_mean <- aggregate(steps ~ interval, data = activity, FUN = mean)

plot(interval_mean, type = 'l', main = 'Average number of steps taken per 5-minute interval')

max_interval <- which.max(interval_mean$steps)
```

The 5-minute interval, on average across all the days in the dataset, containing the maximum number of steps is interval **`r max_interval`**.

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
